Learning words by recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex
reflow all the paragraphs!

% Tell LaTeX to break pages in the middle of displayed math whenever this would be the slightest bit useful
\allowdisplaybreaks[4]

@@

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behaviour, it's useful to begin with optimal models; this has the advantage that when we are then forced to deviate from the purely rational approach, it's generally in a theoretically interpretable way. However, 
Multi-agent interactions are difficult to model in a normative or optimal framework without falling prey to paradox. For example, consider a simple model of the agents in the above game. First we define a \textit{literal listener} $L_0$. This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object\footnote{We assume words refer directly to objects, rather than to abstract semantic features. Our simplification is without loss of generalization, however, because we can interpret our model as marginalizing over such a representation, with our literal $P_\lex(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker should attempt to choose a word which soft-maximizes the probability that the listener will assign to the target object---modulated by the effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not optimal. Instead, listeners should use Bayes rule to invert the speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no longer optimal for speakers to implement strategy $S_1$; instead, they should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

@@

As soon as we attempt to add uncertainty about the meanings of words to such a model, new paradoxes arise. Suppose the listener is a young child who is uncertain about the lexicon their partner is using. The obvious solution is for them to place a prior on the lexicon; they then update their posterior based on whatever utterances and contextual cues they observe, and in the mean time interpret each utterance by making their best guess, marginalizing out this uncertainty. This basic structure is captured in previous models of Bayesian word learning \cite{frank2009}. But when combined with the recursive pragmatic model, a new question arises: Given such a listener, what model should the speaker use? A rational speaker attempts to maximize the listener's likelihood of understanding, so if an uncertain listener interpets by marginalizing over some posterior, then a knowledgeable speaker should calculate this posterior and marginalize themselves. But then the speaker is not using any particular lexicon at all, and a rational learning listener should therefore not be trying to infer the speaker's lexicon.\footnote{Of course, in reality both parties will generally have some uncertainty, making the situation even worse. If we start from an uncertain listener with a prior over lexicons, then a first-level uncertain speaker needs a prior over priors on lexicons, a second-level uncertain listener needs a prior over priors over priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$ recursion was bad enough, but at least each step had a constant cost. This new recursion produces hyper$^n$-distributions for which inference almost immediately becomes intractable even in principle, since the dimensionality of the learning problem increases with each step. Yet, without this addition of new uncertainty at each level, the model would dissolve back into certainty as in the previous paragraph, making learning impossible.}% In fact, such an uncertain listener's uncertainty dissolves; they may not know the lexicon, but they do know their posterior over the lexicon -- and because this is all that the speaker cares about, there is nothing for the listener to learn from the speaker's behaviour.

@@

% . Yet most people do in fact speak languages that they have learned by listening to others, which suggests an additional constraint that should somehow be incorporated into learning: as I build a model of you as a speaker, it should also affect my model of myself as a speaker.

@@

We resolve the problems described above by assuming that speakers and listeners deviate from normative behavior by assuming a conventional lexicon. 
% Clearly we must deviate from pure normativity -- but how? Here, we resolve the above problems by proposing a model of speakers and listeners as rational agents afflicted by a kind of social anxiety. 
Each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't. These assumptions instantiate a kind of ``social anxiety'' in which agents are all trying to learn the correct lexicon that they assume everyone else knows.

Assumption (a) corresponds to the lexicographer's illusion: Naive language users will argue vociferously that words have specific meanings, even though these meanings are unobservable to everyone who purportedly uses them. It also explains why learners speak the language they hear (rather than some private language that they assume listeners will eventually learn): Under assumption (a), observing other speakers' behaviour provides data about not just that speaker's idiosyncratic lexicon, but the consensus lexicon. Assumption (b) avoids the explosion of hyper$^n$-distributions described above: If agent $n$ knows the lexicon, they assume that all lower agents do as well, reducing to the original tractable model without uncertainty. And assumption (c) introduces a limited form of uncertainty, and thus the potential for learning. 
% Everything else follows by applying standard Bayesian reasoning given these assumptions.

@@

% In this section, we describe the coverage of our pragmatic learning model on phenomena previously investigated with the pragmatics and word learning components of the model. To summarize these simulations: the pragmatic learning model reproduces previous findings and provides some new insights into the ``disambiguation'' phenomenon described above.

@@

% XX note the relation to BGL model; they need uncertainty in the middle of the pragmatics specifically, which is where we end up putting it for other reasons. And in Bayesian framework uncertainty and learning must always coincide. Also the intuition that this is a general property of uncertainty for soft-max agents -- if you don't know where you are, good worlds have more peaked utility distributions than bad worlds, and thus more consistent behaviour by other agents, so on average when you're guessing what other agents will do, good worlds contribute more to your average than bad worlds.

@@

% You know, this is almost exactly like a quantitatively graded version of Pascal's wager.

@@

% Even very young children can learn words from individually ambiguous situations, e.g. where multiple words are spoken and multiple objects are present \cite{smith2008}. While no individual situation gives away the mapping, the pattern of co-occurrences between words and objects allows for learning. Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence.

@@

Experimental results suggest that communicators are able to establish novel, consensus-based communication systems. For example, adults playing a communication game using only novel symbols with no conventional meaning will typically converge on a set of new conventions which allow them to accomplish their task \cite{galantucci2005}. Or in a less extreme example, communicators asked to refer to novel objects invent conventional names for them over the course of repeated interactions (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}). From a pure learning perspective this is anomalous, however: Since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour. Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning---even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time. 
% An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).

@@

Fig.~\ref{fig:emergence} shows the result of simulating several such games when both parties begin in total ignorance. Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; and (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons---but they generally recover and converge. Each agent effectively uses their partner's behavior as a basis for forming weak beliefs about the underlying lexicon that they assume must exist. Since they then each act on these beliefs, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a ``superstition'' becomes normatively correct. 
%On the other hand, if the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform. They therefore choose words by simply flipping a coin, and performance remains at chance.

@@

% But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon. The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.

% XXXX what can we say about the role pragmatics plays here? It's annoyingly non-trivial to sample from the equivalent model where the speaker just samples from a multinomial distribution and the listener inverts it. The listener's data is samples from a DM so they have conjugacy. But the speaker's data is listener interpretations, which necessarily do one tiny pragmatic inference step from using Bayes rule, which breaks conjugacy...

>>>>

% Tell LaTeX to break pages in the middle of displayed math whenever
% this would be the slightest bit useful
\allowdisplaybreaks[4]

@@

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behaviour, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each object\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

@@

As soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behaviour.

@@

% . Yet most people do in fact speak languages that they have learned
% by listening to others, which suggests an additional constraint that
% should somehow be incorporated into learning: as I build a model of
% you as a speaker, it should also affect my model of myself as a
% speaker.

@@

We resolve the problems described above by assuming that speakers and
listeners deviate from normative behavior by assuming a conventional
lexicon.
% Clearly we must deviate from pure normativity -- but how? Here, we
% resolve the above problems by proposing a model of speakers and
% listeners as rational agents afflicted by a kind of social anxiety.
Each agent assumes: (a) There is some single, specific literal lexicon
which everyone should be using, (b) and everyone else knows this
lexicon, and believes that I know it as well, (c) but in fact I
don't. These assumptions instantiate a kind of ``social anxiety'' in
which agents are all trying to learn the correct lexicon that they
assume everyone else knows.

Assumption (a) corresponds to the lexicographer's illusion: Naive
language users will argue vociferously that words have specific
meanings, even though these meanings are unobservable to everyone who
purportedly uses them. It also explains why learners speak the
language they hear (rather than some private language that they assume
listeners will eventually learn): Under assumption (a), observing
other speakers' behaviour provides data about not just that speaker's
idiosyncratic lexicon, but the consensus lexicon. Assumption (b)
avoids the explosion of hyper$^n$-distributions described above: If
agent $n$ knows the lexicon, they assume that all lower agents do as
well, reducing to the original tractable model without
uncertainty. And assumption (c) introduces a limited form of
uncertainty, and thus the potential for learning.
% Everything else follows by applying standard Bayesian reasoning
% given these assumptions.

@@

% In this section, we describe the coverage of our pragmatic learning
% model on phenomena previously investigated with the pragmatics and
% word learning components of the model. To summarize these
% simulations: the pragmatic learning model reproduces previous
% findings and provides some new insights into the ``disambiguation''
% phenomenon described above.

@@

% XX note the relation to BGL model; they need uncertainty in the
% middle of the pragmatics specifically, which is where we end up
% putting it for other reasons. And in Bayesian framework uncertainty
% and learning must always coincide. Also the intuition that this is a
% general property of uncertainty for soft-max agents -- if you don't
% know where you are, good worlds have more peaked utility
% distributions than bad worlds, and thus more consistent behaviour by
% other agents, so on average when you're guessing what other agents
% will do, good worlds contribute more to your average than bad
% worlds.

@@

% You know, this is almost exactly like a quantitatively graded
% version of Pascal's wager.

@@

% Even very young children can learn words from individually ambiguous
% situations, e.g. where multiple words are spoken and multiple
% objects are present \cite{smith2008}. While no individual situation
% gives away the mapping, the pattern of co-occurrences between words
% and objects allows for learning. Our current model reproduces this
% pattern of results, converging to correct lexical mappings from
% ambiguous evidence.

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this is anomalous,
however: Since both agents know perfectly well that there is no
underlying convention to discover, there is nothing to learn from the
other's behaviour. Furthermore, even if only one partner is producing
the novel expressions, their behaviour still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behaviour should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (XX
% krauss 1966, branigan catchpole pickering 2011).

@@

Fig.~\ref{fig:emergence} shows the result of simulating several such
games when both parties begin in total ignorance. Notice that: (a)
agents' performance begins at chance, but quickly rises; (b) they tend
towards structured, sparse lexicons with a one-to-one correspondence
between objects and words; and (c) as the speaker and listener have
entirely different data (the listener's interpretations and the
speaker's intended meaning, respectively), unlucky early guesses can
lead them to believe in entirely contradictory lexicons---but they
generally recover and converge. Each agent effectively uses their
partner's behavior as a basis for forming weak beliefs about the
underlying lexicon that they assume must exist. Since they then each
act on these beliefs, and their partner uses the resulting actions to
form new superstitions, they soon converge on using similar lexicons,
and what started as a ``superstition'' becomes normatively correct.
% On the other hand, if the speaker is not allowed to observe the
% listener's guess, then the speaker does no learning and their
% posterior over lexicons remains perfectly uniform. They therefore
% choose words by simply flipping a coin, and performance remains at
% chance.

@@

% But over time, the speaker's use of such patterns, and the
% listener's ability to understand them, will each be taken by the
% other as evidence that these patterns actually exist in the literal
% lexicon. The result is that while both agents start out with uniform
% priors over the lexicon, they are biased to end up believing in
% lexicons that are communicatively useful for the task at hand.

% XXXX what can we say about the role pragmatics plays here? It's
% annoyingly non-trivial to sample from the equivalent model where the
% speaker just samples from a multinomial distribution and the
% listener inverts it. The listener's data is samples from a DM so
% they have conjugacy. But the speaker's data is listener
% interpretations, which necessarily do one tiny pragmatic inference
% step from using Bayes rule, which breaks conjugacy...
